### 2019-06-30 thesis-template/msc-thesis.tex
update

\chapter*{Abstract}
%\input{abstract}
Many problems in science and engineering require
tensor field representations. Tensor fields, are most commonly associated with stress distributions (cf. Cauchy-stress tensor) in 2D/3D grids, but also have some other meanings in practice in physics. By means of global illumination techniques, we motivate a new method to visualize tensor fields, which is also capable of visualizing the light transport (propagation) in a topologically defined scene geometry. As a basis, we derive a simple light propagation scheme for regular grids, which satisfies propagation attenuation and energy-conservation principles and is able to approximate light distributions for given light source position(s) and direction(s) until convergence. The transmission profiles within this grid are considered as crystal fiber structures as an underlying physical model (, eg., gemstones: tiger's eye's cat's eye effect). The consequent task is to model anisotropic fibre structures with the orientation and anisotropy measures of the underlying tensor field in the grid.
For this, we derive an eigenbasis by PCA (principal component analysis) for every cell of the tensor field and form an ellipsoid equation as transmission (transfer/weighting) function for the transmitted light profiles. Hence, the resulting light distribution can be visualized via 2D scalar field color coding or polar plots for directional information. Additionally, we measure impulse responses of the tensor field with Dirac-pulses as light sources at any (sampled) position and in any direction to generate a uniform sampled map of the global illumination distributions. This map is considered as a global illumination energy flow field.
We gain inspiration by vector field particle tracing's FTLE approach for analyzing the gradient of the flow field of the resulting light distributions. Similar to how the FTLE detects ridges/LCS (Lagrangian Coherent Structures) in vector fields (flow fields), our approach captures these structures likewise, namely attracting/repelling/saddling tensor field lines (TFL) which frequently represent key structures in nature, in tensor fields. We denote this entity a light transport gradient (LTG) which states an FTLE-related approach for visualizing LCS in tensor fields through analysis of the global light transport  directed by imprint/engraving in crystelline fiber structures modeled by the tensor field's ellipsoid glyhps. We evaluate our approaches for plausibility and comparability within an extensive test campaign and demonstrate its performance on several examples of datasets.
\newpage

@@

\paragraph{Contributions of this Work}
\begin{itemize}
	\item light transport model (propagation scheme) following basic but crucial physical principles for tensor field visualization using tensors as transmission profiles
	\item visualization of the propagation results in $2D$ polar profiles and scalar fields
	\item a FTLE-related approach called light transport gradient (LTG) for visualizing key structures (ROIs) namely LCS in $2D$ second-order tensor fields
	\item application of our approach to real examples featuring brain and heart datasets.
\end{itemize}
%Dieses Kapitel gibt einen Überblick über die Arbeit. Gerade der
%Abschnitt zur Motivation soll allgemein verständlich geschrieben
%werden. Die Einleitung sollte auch wichtige Referenzen enthalten.

\section{Motivation}
\label{sect:motiv}
Tensor field visualization is gaining importance as a relevant tool for the analysis of fluid and solid mechanics, where tensor representations are crucially needed. This is always the case, when a directional distribution is needed for a single point in space. Tensors occur most commonly in medical, scientific and engineering applications, making them a frequently used mathematical entity in science. In general, a tensor is a $o$-D generalization of scalar, vector and matrix and appears, e.g., as Jacobian matrix (spatial gradient) in vector (flow) field visualization or as stress tensor in solid and fluid continuum mechanics. Basically, tensors describe compressions, tensions, rotations and volume changes in both solid and fluid material. While most techniques focus on symmetric tensor field visualization (like glyphs and tensor field lines), some other recent works adress the problem of asymmetric tensor field visualization. The issue with asymmetric (non-symmetric) or included antisymmetric (skew-symmetric) tensors is, that they do not always yield real eigenvalues, which is needed to determine an eigenbasis to set up glyphs, hyperstreamlines or tensor field lines (classically). Nevertheless, there has been extensive work in this domain as well. Zheng \& Phang et al. \cite{pang&zheng} proposed the concept of dual eigenvectors and Lin et al. \cite{zhang} extended it to pseudo-eigenvectors and introduced eigenvector and eigenvalue manifold to visualize eigenvectors in the complex domain. In this work, we will use the $\mathbf{U}$-matrix obtained from singular value decomposition (SVD) to obtain the half-axes of the PCA principal ellipsoid representing the tensor transformation as suggested by Moler \cite{moler} to state another alternative of processing asymmetric tensors. 2D tensor field visualization is useful for cases where only two-dimensional data is available, such as satellite remote sensing data, Particle Imagery Velocimetry (PIV) data and earthquake data \cite{laramee}. Additionally, tensor field visualization aids geologists and medical scientists in their field of work to display complicated number arrays which seem almost impossible to interpret in their raw, unprocessed form. Nevertheless, comprehensive visualization is aiding scientists in their everyday lifes understanding complex data loads such as tensor fields. To cope with the arising problem of tensor field visualization, we can only try to invent new methods continuously evaluating them for suitability for certain domains of science to provide our techniques at best considering them for the right applications.\\

@@

  \label{Skalierung}

@@

The main purpose of visualization in general is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor-medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue or in simulated (or measured) stress tensor fields from solid and fluid mechanics. In these domains of science the problem arises, that tensor fields can be generated (simulated) or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e., mostly 2D/3D-grids consisting of mainly second-order tensors (cf. Cauchy stress tensor). These grids span a spatial domain in general with sample positions indicated by vertices. Tensor fields feature a sampling grid consistent of tensors which is spaced uniformly in the commonly used regular grid. A regular grid, which we will utilize in this context, is a normalized uniform grid as a special case of a regular grid. We aim to use global illumination methods to model light transport directed by tensor fields within such type of grids, which we consider as light propagation volumes (LPVs). Congruently, the most basic objective of this work is to design a simple light propagation scheme yet following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, for tensor field visualization. Furthermore, it should hold the additional possibility to model light transport in a geometrical scene. This propagation scheme should either be used to visualize the light transport paths (,defined by the anisotropy of tensor fields and tensor magnitudes of a given tensor field in a $2D$ grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a $2D$ grid. However, the tensor field needs to be pre-processed somehow to generate a tensor induced footprint map of transmission profiles, which we will do via matrix decompositions. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of tensor controlled directive effect on the light transport. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (eg. $r(\omega)=1$) by the user which we sample as a primary step. This way, the approach will operate as a global illumination visualization technique for light distributions in polar coordinates provided the user set an initial profile into the grid. In the sense of automization and optimzation, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to graps by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (forming a saddle) local features, which frequently represent key structures in nature such as moraines in glaciers. LCS usually separate regions of different flow behavior, but in our case they separate regions/domains of varying directional distributions (manifesting in diverging/converging tensor field lines or higher tensor magnitude) represented by tensors as we use our tensor controlled light transport method as a basis or framework to compute the FTLE field. Thus, we propose another method to compute an FTLE like field on tensor fields. The FTLE map responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other because of bidirectional stress directions and since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain elaborate representations and the interpretation of meaningful results comparing the proposed techniques against others. The designed approach could e.g. be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations.

@@

For the rest of this work, we will give a quick overview on the structure and outline. Next, in chap. \ref{chap:grundlagen}, we will capture important fundamentals like Tensors, polar coordinates, PCA in the context of singular value decomposition (generalization of eigenvalue decomposition) and GT datasets. Also we will discuss the most relevant related work in the domain of tensor field visualization. In chap. \ref{chap:main}, we will propose the main contributions and methods of this work and document the results and evaluation in chap. \ref{chap:eval}. Subsequently, A compact summary, conclusion and outlook will be given in chap. \ref{chap:concl}.\\

@@

In this chapter, we will capture the prerequisites necessary to understand the contributions of this work and we will capture tensor fields, polar coordinates, PCA, GT datasets and related work in that sense.

@@

\section{Tensor Fields}
\label{sec:tensors}
Now what is a tensor? A tensor is a $o$-dimensional generalization of a matrix as illustrated in Table \ref{tensor}.
\begin{table}[!b]
\centering
\caption{Tensor Shapes}
 \begin{tabular}{c|c|c|c|c|c|c}
\textbf{order} & 0 & 1 & 2 & 3 & ... & o \\ 
\hline 
\textbf{shape} & scalar & vector & matrix & ``3D matrix'' & ... & $\mathop{o}$D matrix\\ 
\end{tabular}
\label{tensor}
 \end{table} 
In addition to that, a tensor has as many indices as its order $o$ and their run length is as long as its embedded dimension $n$ in space (the dimension of the row and column vectors), which is equal to the rank of the tensor for full rank (in other words the number of elements n for a $n{\times}n$-matrix). Tensors follow certain transformation rules which are defined for covariant/contravariant tensors, which indicate the change characteristics under a basis change. Contravariant tensors do not change under a change of basis, whereas covariant tensors are not invariant w.r.t. basis changes. Tensors are most commonly obtained and defined from mathematical transformation equations. Hence, any $oD$ number array could be a tensor, but the definition holds only if the transformation rules apply to these $oD$ number arrays (in a sense of mathematical/formal interpretation). This is mostly true for component indexed $oD$ number entities following matrix multiplication and transformation rules found in physics and math manifesting in calculation rules like matrix multiplication. Thus, a tensor in general is a $o$D number array arranged in meaningful (transformation) order to put it simple. In fact, tensors themselves are interpreted as transformations which reflect an incoming direction vector (measuring, e.g., the stress in that direction) to an outgoing (stress) vector. To be a bit more concrete, tensors represent stress distributions in solids and fluids describing strain and flow features as these entities are not simple numbers or vectors. They occur in form of a tensor fields, and thus as entities or quantities in physics. That is, at each point (typically a Euclidean space or manifold) in space (indexed by the grid) there is a whole distribution of directions, e.g., stresses (elastic and viscous), which needs to be characterized and represented by a single tensor each. A tensor is considered non-degenerate if all eigenvalues $\lambda_i>\lambda_{i+1}$ manifesting in significant anisotropy.
 \\
 \paragraph{stress tensor}
 \begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.6\textwidth]
    {img/stress_tensor.png}
  }
  \caption{Cauchy stress Tensor}
  \captionsetup{font={footnotesize,bf,it}}
  \caption*{src: \url{https://en.wikipedia.org/wiki/Cauchy_stress_tensor\#/media/File:Components_stress_tensor_cartesian.svg}}
  \label{cauchy}
\end{figure}

The Cauchy stress tensor, which is propably the classical example of a tensor, is depicted in Fig. \ref{cauchy} and consists of 3 stress vectors $\mathbf{T(e_i)}$ arranged in row-major order. It appears in similar form as viscous stress tensor in fluid mechanics:
\begin{align*}
\boldsymbol{\sigma}
 = \left[{\begin{matrix} \mathbf{T}^{(\mathbf{e}_1)} \\
\boldsymbol{T}^{(\mathbf{e}_2)} \\
\boldsymbol{T}^{(\mathbf{e}_3)} \\
\end{matrix}}\right] =
\left[{\begin{matrix}
\sigma _{xx} & \sigma _{xy} & \sigma _{xz} \\
\sigma _{yx} & \sigma _{yy} & \sigma _{yz} \\
\sigma _{zx} & \sigma _{zy} & \sigma _{zz} \\
\end{matrix}}\right]  = \left[{\begin{matrix}
\sigma _x & \tau _{xy} & \tau _{xz} \\
\tau _{yx} & \sigma _y & \tau _{yz} \\
\tau _{zx} & \tau _{zy} & \sigma _z \\
\end{matrix}}\right] =
\begin{bmatrix}
\sigma _{11} & \sigma _{12} & \sigma _{13} \\
\sigma _{21} & \sigma _{22} & \sigma _{23} \\
\sigma _{31} & \sigma _{32} & \sigma _{33} 
\end{bmatrix}.
\end{align*}
These 3 stress vectors represent the orientation and magnitude of total resulting stress at plane ${x,y,z}$ in direction ${x,y,z}$ ($3^2=9$ individual numbers). Thus, the stress tensor poses a full valid represenation of the stress distribution at any point in space. The stress tensor itself can be interpreted as a transformation which maps an incoming normal direction vector $\mathbf{n}$ a resulting stress vector $\mathbf{T^{(n)}} = \mathbf{n}\cdot\boldsymbol{\sigma}$. The stresses in normal directions (diagonal elements) are directly related to the pressure, which is always isotropic and hence non-directionally dependent (kind of a Euclidean distance of the stress components). The tensor itself is a complicated representation and needs to be projected onto its principal axes to lower its dimensionality. It is possible to find 2 principal stress directions in 2D through PCA (see sect. \ref{sec:pca} below), for which there are no shear stresses existent in equilibrium. These form the principal directions of deformation at a certain location and are sufficient to describe the resulting transformation/deformation behaviour. The resulting system of principal directions and absolute stresses is often called an ``eigensystem'' or ``eigenframe''. So to say, an eigensystem, consistent of the principal directions/axes, spans a principal ellipsoid depicted in Fig. \ref{hat} in sect. \ref{sec:pca}.
Another interpretation for the principal component ellipsoid is that it is the result of the tensor transformation applied to the unit circle. It is therefore somehow representing the distortion of an absolutely isotropic object (unit circle) by whatever the affine tensor transformation effects (e.g. stretching, shearing, mirroring, scaling, rotating). Let us consider it as some lower dimensional, more intuitive representation of the tensor which we were left with in the first place. To put it all together, a tensor is a mathematical transformation which is encoded inside of an abstract $oD$ number array, which comes in need, whenever we have a whole distribution of directions manifesting itself in a transformation ellipsoid.

\section{Polar Coordinates}
\label{sec:polar}
\begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.8\textwidth]
    {img/polarkoordinaten.png}
  }
  \caption{polar space}
  \captionsetup{font={footnotesize,bf,it}}
   \caption*{src: \url{https://de.wikipedia.org/wiki/Datei:Ebene_polarkoordinaten.PNG}}
  \label{polar}
\end{figure}\vskip 3pt
Mathematically speaking, a polar space is a system which maps each angle $\omega$ a magnitude $r(\omega)$ through a symbolic function. The conversion formula from polar (given a radius $r$ and angle $\omega$) to Cartesian coordinates is depicted in Fig. \ref{polar} and is given by:
\begin{align*}
	x &= r \cos \omega, \\
 	y &= r \sin \omega.
\end{align*}
Note that we switched angle $\phi$ to $\omega$ for reasonable discrimination to the radiant intensity. The inverse direction (given coordinates $x,y$) is a bit more complicated:
\begin{align*}
	r &= \sqrt{x^2 + y^2} \\
\omega &= \operatorname{atan2}(y, x)
\end{align*}
The polar domain is a circular axis version of the domain $[0..2\pi]$, with negative magnitudes in $r$ shifted by 180 degrees (point reflected) as this is the formal behaviour defined and required by the conversion formulas.

\section{Principal Component Analysis}
\label{sec:pca}
Note that we denote a matrix with a second order tensor interchangeably for the sake of simplicity in the following. Also keep in mind, that we interpret a matrix or tensor as affine transformation. For any state of stress in equilibrium it is possible to find $n$ (dimensionality) independent orthogonal directions with no shear stresses existent. Along these directions the principal stresses, which are normal (tensile/compressive) stresses, are exerted. The anisotropy can be represented by, e.g., ellipsoid glyphs.

@@

\begin{figure}[!t]
\centering
  \begin{minipage}{0.35\textwidth}
    \includegraphics[width=\textwidth]{img/drain.png}
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.35\textwidth}
    \includegraphics[width=\textwidth]{img/drain-TFL.png}
    \label{b)}
  \end{minipage}
\caption{a) ``drain''-testfield, b) tensor field lines for a)}
\label{drain}
\end{figure}

The ``drain''-Testfield, depicted in Fig. \ref{drain} is generated to have a kind of proof-of-concept (POC) for the FTLE-related approach. It poses raw, bare diverging tensor field lines without any other feature modulated on top. Thus, it can be expected to induce a high FTLE response, since it is an interpretable ground truth test and is considered to be a basic trigger/stimulus for the approach.

>>>>

\chapter*{Abstract}
%\input{abstract}
Many problems in science and engineering require
tensor field representations. Tensor fields, are most commonly associated with stress distributions (cf. Cauchy-stress tensor) in 2D/3D grids, but also have some other meanings in practice in physics. By means of global illumination techniques, we motivate a new method to visualize tensor fields, which is also capable of visualizing the light transport (propagation) in a topologically defined scene geometry. As a basis, we derive a simple light propagation scheme for Cartesian grids, which satisfies propagation attenuation and energy-conservation principles and is able to approximate light distributions for given light source position(s) and direction(s) until convergence. The transmission profiles within this grid are considered as crystal fiber structures as an underlying physical model (, e.g., gemstones: tiger's eye's cat's eye effect). The consequent task is to model anisotropic fiber structures with the orientation and anisotropy measures of the underlying tensor field in the grid. For this, we derive an eigenbasis by PCA (principal component analysis) for every cell of the tensor field and form an ellipsoid equation as transmission (transfer) function for the transmitted light profiles. Hence, the resulting light distribution can be visualized via 2D scalar field color coding or polar plots for directional information. On this basis, we measure impulse responses of the tensor field with Dirac-pulses as light sources at any (sampled) position and in any direction to generate a uniform sampled map of the global illumination distributions. This map is considered as a global illumination energy flow field. We gain inspiration by vector field particle tracing's FTLE approach for analyzing the gradient of the flow field, formed by the resulting light distributions. Similar to how the FTLE detects ridges/LCS (Lagrangian Coherent Structures) in vector fields (flow fields), our approach captures these structures likewise, namely regions revealing attracting/repelling/saddling tensor field lines (TFL) which frequently represent key structures in nature, in tensor fields. We denote this entity as light transport gradient (LTG) which states an FTLE-related approach for visualizing LCS in tensor fields through analysis of the global light transport in the grid, directed by imprints in crystelline fiber structures modeled by the tensor field's ellipsoid glyhps. We evaluate our approaches for plausibility and comparability within an extensive test campaign and demonstrate its performance on several examples of both synthetic and real datasets.
\newpage

@@

\paragraph{Contributions of this Work}
\begin{itemize}
	\item light transport model (propagation scheme) following basic but crucial physical principles for tensor field visualization using tensors as transmission profiles,
	\item visualization of the propagation results in $2D$ polar profiles and scalar fields,
	\item a FTLE-related approach called light transport gradient (LTG) for visualizing key structures (ROIs), namely LCS in $2D$ second-order tensor fields, and
	\item application of our approach to real data involving brain and heart datasets.
\end{itemize}
%Dieses Kapitel gibt einen Überblick über die Arbeit. Gerade der
%Abschnitt zur Motivation soll allgemein verständlich geschrieben
%werden. Die Einleitung sollte auch wichtige Referenzen enthalten.

\section{Motivation}
\label{sec:motiv}
Tensor field visualization is gaining importance as a relevant tool for the analysis of fluid and solid mechanics, where tensor representations are crucially needed. This is always the case, when a directional distribution is needed for each single point in space. Tensors occur most commonly in medical, scientific and engineering applications, making them a frequently used mathematical entity in science. In general, a tensor is a $o$-D generalization of scalar, vector and matrix and appears, e.g., as Jacobian matrix (spatial gradient) in vector (flow) field visualization or as stress tensor in solid and fluid continuum mechanics. Basically, tensors describe compressions, tensions, rotations and volume changes in both solid and fluid material, as illustrated in Fig. \ref{ToS}. While most techniques focus on symmetric tensor field visualization (like glyphs and tensor field lines), some other recent works adress the problem of asymmetric tensor field visualization. The issue with asymmetric (non-symmetric) or therein included antisymmetric (skew-symmetric) tensors is, that they do not always yield real eigenvalues, which is needed to determine an eigenbasis to set up glyphs, hyperstreamlines or tensor field lines (classically). Nevertheless, there has been extensive work in this domain as well. Classically, the issue with asymmetric tensor fields is, that they do not yield real eigenvalues in general, which is needed to set up glyph representations. In this work, we will use the $\mathbf{U}$-matrix obtained from singular value decomposition (SVD) to obtain the half-axes of the PCA principal ellipsoid representing the tensor transformation as suggested by Moler \cite{moler} to state another alternative of processing asymmetric tensors. 2D tensor field visualization is useful for cases where only two-dimensional data is available, such as satellite remote sensing, particle imagery velocimetry (PIV) and earthquake data \cite{laramee}. In their field of work, geologists and medical scientists are struggling with complicated number arrays which seem almost impossible to interpret in their raw, unprocessed form. Nevertheless, comprehensive tensor field visualization is aiding scientists in their everyday lifes understanding complex data loads, such as tensor fields. To cope with the arising problem of tensor field visualization, we can only try to invent new methods, continuously evaluating them for suitability for certain domains of science, to provide our techniques at best, while considering them for the right applications.\\

@@

  \label{streamlines}

@@

The main purpose of visualization in general, is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor - medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue or in simulated (or measured) stress tensor fields from solid and fluid mechanics. In these named domains of science the problem arises, that tensor fields can be generated (simulated) or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e., mostly 2D/3D-grids, commonly consisting of second-order tensors (cf. Cauchy stress tensor). In general, these grids span a spatial domain with sample positions indicated by vertices or cell centers in the dual grid, depending on the mode of interpretation. Tensor fields feature a sampling grid, consistent of tensors, which is spaced uniformly in the commonly used regular grid. A Cartesian grid, which we will utilize in this context, is a grid normalized to a certain unit length, as a special case of a regular grid. That is, each cell is formed as a unit cube in relation to the normalizing unit length. We aim to use global illumination methods to model light transport, directed by tensor fields, within such type of grids, which we consider as light propagation volumes (LPVs) \cite{dachsbacher}. Congruently, the most basic objective of this work is to design a simple light propagation scheme following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, for tensor field visualization. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (, e.g., $r(\omega)=1$) by the user which we sample (discretize) as a primary step. This way, the approach will operate as a light transport visualization technique for light distributions given in polar coordinates, provided the user set an initial profile into the grid. Furthermore, this propagation scheme should either be used to visualize the light transport paths, defined by the anisotropy characteristics of tensor fields and tensor magnitudes of a given tensor field in a $2D$ grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a $2D$ grid. However, the tensor field needs to be preprocessed somehow to generate a footprint map of transmission profiles defined by the tensor field defined by the tensor field's anisotropy characteristics, which we will handle through matrix decompositions. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of tensor controlled directive effect on the light transport. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. This will yield an approach, capable of computing approximated final light distributions for a given light source configuration inside of a $2D$ lattice. As this process requires a lot of user-interaction and is not to be concerned as generalizing to all possible scenarios. In the sense of automization and optimzation, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to grasp by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (forming a saddle) local features, which frequently represent key structures in nature, such as moraines in glaciers or vortices in flow fields. Commonly, in vector field visualization, LCS separate regions of different flow behavior, but in our case they separate regions/subdomains of varying directional distributions (manifesting in diverging/converging tensor field lines $\rightarrow$ compressions, tensions, roations or varying tensor magnitude $\rightarrow$ volume changes) represented by tensors, as we use our tensor-controlled light transport visualization (propagation scheme) as a basis or framework to compute the FTLE field. Consequently, we propose another method to compute an FTLE-like field on tensor fields. The FTLE field responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other because of bidirectional stress directions and since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide new tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain abstracted representations and the interpretation of meaningful results evaluating the proposed techniques for possible uses. The designed approach could ,e.g. , be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations. A classical example for tensor field visualization is plotting hyperstreamlines following major/minor eigenvectors, as depicted in Fig. \ref{streamlines}, which features a degenerate point (purple), a wedge (red) and a trisector (green).

@@

For the rest of this work, we will give a quick overview on the structure and outline. Next, in chap. \ref{chap:grundlagen}, we will capture important fundamentals like tensors, polar coordinates, PCA in the context of singular value decomposition (generalization of eigenvalue decomposition), FTLE and GT datasets. Also we will discuss the most relevant related work in the domain of tensor field visualization and also in the domain of global illumination (indirect illumination rendering). Subsequently, in chap. \ref{chap:main}, we will propose the main contributions and methods of this work and document the results and evaluation in chap. \ref{chap:eval}. At last, a compact summary, conclusion and outlook will be given in chap. \ref{chap:concl}.\\

@@

In this chapter, we will capture the prerequisites necessary to understand the contributions of this work and we will capture tensor fields, polar coordinates, PCA, FTLE, GT datasets and related work in that sense.

@@

\section{Tensor Fields}
\label{sec:tensors}
Now what is a tensor? A tensor is a $o$-dimensional generalization of a matrix as illustrated in Table \ref{tensor}.
\begin{table}[!b]
\centering
\caption{Tensor Shapes}
 \begin{tabular}{c|c|c|c|c|c|c}
\textbf{order} & 0 & 1 & 2 & 3 & ... & o \\ 
\hline 
\textbf{shape} & scalar & vector & matrix & ``3D matrix'' & ... & $\mathop{o}$D matrix\\ 
\end{tabular}
\label{tensor}
 \end{table} 
In addition to that, a tensor has as many indices as its order $o$ and their run length is as long as its embedded dimension $n$ in space (the dimension of the row and column vectors), which is equal to the rank of the tensor for full rank (in other words the number of column elements n for a $n{\times}n$-matrix). A tensor is considered non-degenerate if all eigenvalues $\lambda_i>\lambda_{i+1}$ manifesting in significant anisotropy. Tensors follow certain transformation rules which are defined for covariant/contravariant tensors, which indicate the change characteristics under a basis change. Contravariant tensors do not change under a change of basis, whereas covariant tensors are not invariant w.r.t. basis changes. Tensors are most commonly obtained and defined from mathematical transformation equations. Hence, any $oD$ number array could be a tensor, but the definition holds only if the transformation rules apply to these $oD$ number arrays (in a sense of mathematical/formal interpretation). This is mostly true for component indexed $oD$ number arrays following matrix multiplication and transformation rules found in physics and math manifesting in calculation rules like matrix multiplication. Thus, a tensor in general is a $o$D number array arranged in meaningful (transformation) order, to put it simple. In fact, tensors themselves are interpreted as transformations which reflect an incoming direction vector (measuring, e.g., the stress in that direction) to an outgoing (stress) vector. To be a bit more concrete, tensors represent stress distributions in solids and fluids describing strain and flow features as these entities are not simple numbers or vectors. They occur in form of a tensors, and thus as entities or quantities in physics. That is, at each point (typically a Euclidean space or manifold) in space (indexed by the grid) there is a whole distribution of directions, e.g., stresses (elastic and viscous), which needs to be characterized and represented by a single tensor each. Collectively, defined at each point in space, these tensors form a tensor field.
 \paragraph{stress tensor}
 \begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.6\textwidth]
    {img/stress_tensor.png}
  }
  \caption{Cauchy stress Tensor}
  \captionsetup{font={footnotesize,bf,it}}
  \caption*{src: \url{https://en.wikipedia.org/wiki/Cauchy_stress_tensor\#/media/File:Components_stress_tensor_cartesian.svg}}
  \label{cauchy}
\end{figure}

The Cauchy stress tensor, which is propably the classical example of a tensor, is depicted in Fig. \ref{cauchy} and consists of 3 stress vectors $\mathbf{T(e_i)}$, arranged in row-major order. It appears in similar form as viscous stress tensor in fluid mechanics:
\begin{align*}
\boldsymbol{\sigma}
 = \left[{\begin{matrix} \mathbf{T}^{(\mathbf{e}_1)} \\
\boldsymbol{T}^{(\mathbf{e}_2)} \\
\boldsymbol{T}^{(\mathbf{e}_3)} \\
\end{matrix}}\right] =
\left[{\begin{matrix}
\sigma _{xx} & \sigma _{xy} & \sigma _{xz} \\
\sigma _{yx} & \sigma _{yy} & \sigma _{yz} \\
\sigma _{zx} & \sigma _{zy} & \sigma _{zz} \\
\end{matrix}}\right]  = \left[{\begin{matrix}
\sigma _x & \tau _{xy} & \tau _{xz} \\
\tau _{yx} & \sigma _y & \tau _{yz} \\
\tau _{zx} & \tau _{zy} & \sigma _z \\
\end{matrix}}\right] =
\begin{bmatrix}
\sigma _{11} & \sigma _{12} & \sigma _{13} \\
\sigma _{21} & \sigma _{22} & \sigma _{23} \\
\sigma _{31} & \sigma _{32} & \sigma _{33} 
\end{bmatrix}.
\end{align*}
These 3 stress vectors represent the orientation and magnitude of total resulting stress at plane ${x,y,z}$ in plane normal directions ${x,y,z}$ ($3^2=9$ individual numbers). Thus, the stress tensor poses a full valid represenation of the stress distribution at any point in space. The stress tensor itself can be interpreted as a transformation, which maps an incoming normal direction vector $\mathbf{n}$ a resulting stress vector $\mathbf{T^{(n)}} = \mathbf{n}\cdot\boldsymbol{\sigma}$. The stresses in normal directions (diagonal elements) are directly related to the pressure, which is always isotropic and hence non-directionally dependent (kind of a Euclidean distance of the stress components). The tensor itself is a complicated representation and needs to be projected onto its principal axes to lower its dimensionality, as it is the case for principal component analysis. It is possible to find $2$ principal stress directions in $2D$ through PCA (see sect. \ref{sec:pca} below), for which there are no shear stresses existent in equilibrium. These form the principal directions of deformation at a certain location and are sufficient to describe the resulting transformation/deformation behaviour. The resulting system of principal directions and absolute stresses is often called an ``eigensystem'' or ``eigenframe''. So to say, an eigensystem, consistent of the principal directions/axes, spans a principal ellipsoid depicted in Fig. \ref{hat} in sect. \ref{sec:pca}. Another interpretation for the principal component ellipsoid is that it is the result of the tensor transformation applied to the unit circle. It is therefore somehow representing the distortion of an absolutely isotropic object (unit circle) by whatever the underlying affine tensor transformation effects (e.g. stretching, shearing, mirroring, scaling, rotating). Let us consider it as some lower dimensional, more intuitive representation of the tensor which we were left with in the first place. To put it all together, a tensor is a mathematical transformation which is encoded inside of an abstract $oD$ number array, which comes in need, whenever we have a whole distribution of directions manifesting itself inside of a tensor (matrix) transformation.

\section{Polar Coordinates}
\label{sec:polar}
\begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.8\textwidth]
    {img/polarkoordinaten.png}
  }
  \caption{polar space}
  \captionsetup{font={footnotesize,bf,it}}
   \caption*{src: \url{https://de.wikipedia.org/wiki/Datei:Ebene_polarkoordinaten.PNG}}
  \label{polar}
\end{figure}\vskip 3pt
Mathematically speaking, a polar space is a system, which maps each angle $\omega$ a magnitude $r(\omega)$ through a symbolic function. The conversion formula from polar (given a radius $r$ and angle $\omega$) to Cartesian coordinates is depicted in Fig. \ref{polar} and is given by:
\begin{align*}
	x &= r \cos \omega, \\
 	y &= r \sin \omega.
\end{align*}
Note, that we switched angle $\phi$ to $\omega$ for reasonable discrimination to the radiant intensity. The inverse direction (given coordinates $x,y$) is a bit more complicated:
\begin{align*}
	r &= \sqrt{x^2 + y^2} \\
\omega &= \operatorname{atan2}(y, x)
\end{align*}
The polar domain is a circular axis version of the domain $[0..2\pi]$, with negative magnitudes in $r$ shifted by 180 degrees (point reflected) as this is the formal behaviour defined and required by the conversion formulas.

\section{Principal Component Analysis}
\label{sec:pca}
Note, that we denote a matrix with a second order tensor interchangeably for the sake of simplicity in the following. Also keep in mind, that we interpret a matrix or tensor as affine transformation. For any state of stress in equilibrium it is possible to find $n$ (dimensionality) independent orthogonal directions with no shear stresses existent. Along these directions the principal stresses, which are normal (tensile/compressive) stresses, are exerted. The anisotropy can be represented by, e.g., ellipsoid glyphs.

@@

\begin{figure}[!t]
\centering
  \begin{minipage}{0.35\textwidth}
    \includegraphics[width=\textwidth]{img/drain(alt).png}
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.35\textwidth}
    \includegraphics[width=\textwidth]{img/drain(alt)-TFL.png}
    \label{b)}
  \end{minipage}
\caption{a) ``drain''-testfield, b) tensor field lines for a)}
\label{drain}
\end{figure}

The ``drain''-Testfield, depicted in Fig. \ref{drain} is generated to have a kind of proof-of-concept (POC) for the FTLE-related approach. It poses raw, bare diverging tensor field lines without any other feature modulated on top. Thus, it can be expected to induce a high FTLE response, since it is an interpretable ground truth test and is considered to be a basic trigger/stimulus for the approach with no side effects introduced.
